### 2014-09-16 paper.tex
minor update
misc fixes

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Splice variants play an important role in biological systems, especially in the
immune system and the central nervous system. An expression profile of splice
variants has been shown to be a better signature of some diseases than an
overall gene expression profile~\cite{zhang2013isoform}. However, for most
organisms, available gene models such as Ensembl may not include all splice
variants expressed in RNA-Seq data. Reference-guided (e.g.
Cufflinks~\cite{Trapnell:2010kd}) and de novo assembly (e.g.
Velvet/Oases~\cite{Schulz:2012je}) methods are available for building gene
models from RNA-Seq reads. However, we found that both types of method recovered
distinct sets of splice variants.  In this article, we introduce a pipeline that
can combine transcripts from {\em de novo} assembly and other gene models to
constructs gene models that include more splice variants. We also described a
method called local assembly that we used to enhance the assembly of splice
variants to recover splice variants not found by Cufflinks and {\em de novo}
assembly.  Combination of Cufflinks gene models and transcripts from assembly
include more splice variants, increasing sensitivity of alternative splicing
study.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

@@

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to high-cost and low-throughput
sequencing of expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into a complexity of transcriptomes.

It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately, $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.

Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

Several tools have been developed not only to detect novel splice
junctions but also to reconstruct full-length isoforms from short
reads without using prior gene annotations.  These tools are especially
useful for transcriptome analysis of organims with incomplete gene
annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
that can align reads across putative exon junctions, to reconstruct a
full-length transcript.  Cufflinks~\cite{} identified 12,712 novel isoforms,
of which 7,395 ($58\%$) contain novel splice junctions in mouse
myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
employing a similar mapping-based approach, to reconstruct a full-length
transcripts from mouse RNA-Seq data and discovered approximately 490
novel alternative isoforms in lincRNA loci, which are expressed in any
of the three different cell types\cite{Guttman:2010io}.

Although these mapping-based methods have been useful in detecting
both splice junctions and isoforms, they rely heavily on a reference
genome.  Hence, it is not necessarily practical to apply these methods
to organisms lacking a high-quality reference genome.  This limitation
can be overcome by {\em de novo} assembly of short reads.

A number of {\em de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb} was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
approach is capable of increasing sensitivity of detecting
alternative isoforms over a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior to other {\em de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

@@

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly (Table~\ref{unique_sequences},

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank (Fig.~\ref{chick_est_venn}).  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

@@

For organisms with a reference genome and Ensembl annotation, building gene
models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
reference guide seems to be a preferable method for gene and isoform expression
analysis.  However, we have shown that Ensembl models in chicken are missing a
substantial number of splice variants found in ESTs and mRNAs and Cufflinks does
not recover all splice variants.  A previous study by Schulz \emph{et
al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
assembly in detecting genes and isoforms in Ensembl annotations in mouse.
However, we found that Cufflinks almost exclusively finds splice junctions in
mouse Ensembl gene models whereas our gene models include splice junctions not in
Ensembl gene models but supported by ESTs (Figure~\ref{mus_venn}).

To build mouse Cufflinks gene models, we used Tophat default parameters claimed
to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
detected splice junctions not in Ensembl.  This suggests that the parameters may
have been set based on mammalian Ensembl models, which could limit the
efficiency of finding novel splice variants in other non-mammal organisms.  In
contrast, {\em de novo} assembly does not rely on species-specific parameters.
It is, therefore, recommended that a combination of Cufflinks models and
transcripts from {\em de novo assembly} should be used to build gene models that
include more splice variants.

@@

Both Cufflinks and our pipeline rely heavily on a reference genome; therefore,
the quality of the genome will greatly affect the quality of the gene models.
In this study gene models were built from chicken genome version 2.1 (galGal3),
which contains ~17 Mb of sequence duplications and missassemblies
that were eliminated in the latest version of genome assembly
(galGal4).  Duplications and misassemblies lead to false splice junctions, which
in turn produce a large number of splice variants as observed in some
chromosomes.

@@

\subsection*{Data}
Mouse RNA-Seq dataset (SRX062280) is downloaded from Short Read Archives (SRA).
Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from spleen of
chicken line 6 and 7.

@@

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons one of whose splice sites perfectly match intron boundaries.
Exons are considered incomplete and eliminated if they locate at the $3'$ or
$5'$ end of the transcripts and they are not the largest exons.  ,
Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
they have at least one intron or exon in common.  Then, a splice graph composed
of exons is created and structures of isoforms are derived from traversing
paths in the splice graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

To identify unique sequences from two datasets, a set of 20-mers is created for
both datasets using khmer\cite{khmer}.  Then, 20-mers from a query dataset are
compared with 20-mers from the target dataset.  The sequence is considered
unique if more than 90\% of 20-mers in the query is unique.  Any unique region
shorter than 300 bp is ignored.

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Splice variants play an important role in biological systems, especially in the
immune system and the central nervous system. An expression profile of splice
variants has been shown to be a better signature of some diseases than an
overall gene expression profile~\cite{zhang2013isoform}. However, for most
organisms, available gene models such as Ensembl may not include all splice
variants expressed in RNA-Seq data. Reference-guided (e.g.
Cufflinks~\cite{Trapnell:2010kd}) and de novo assembly (e.g.
Velvet/Oases~\cite{Schulz:2012je}) methods are available for building gene
models from RNA-Seq reads. However, we found that both types of method recovered
distinct sets of splice variants.  In this article, we introduce a pipeline that
can combine transcripts from {\em de novo} assembly and other gene models to
constructs gene models that include more splice variants. We also described a
method called local assembly that we used to enhance the assembly of splice
variants to recover splice variants not found by Cufflinks and {\em de novo}
assembly.  Combining Cufflinks gene models and transcripts from assembly
yields more splice variants, increasing sensitivity of alternative splicing
studies.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

@@

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to the high cost and low throughput
of sequencing expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into the complexity of transcriptomes (@cite).
It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.
Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

Several reference-based tools have been developed not only to detect
novel splice junctions but also to reconstruct full-length isoforms
from short reads without using prior gene annotations.  These tools
are especially useful for transcriptome analysis of organims with
incomplete gene annotations.  Cufflinks\cite{Trapnell:2010kd} relies
on splice junctions detected from Tophat\cite{Trapnell:2009dp}, a read
aligner that can align reads across putative exon junctions, to
reconstruct a full-length transcript.  In mRNAseq from mouse myoblast
cell lines, Cufflinks identified 12,712 novel isoforms, of which 7,395
($58\%$) contain novel splice junctions.  Guttman {\em et al} used
Scripture, a tool employing a similar mapping-based approach, to
reconstruct full-length transcripts from mouse RNA-Seq data and
discovered approximately 490 novel alternative isoforms in lincRNA
loci, which are expressed in any of the three different cell
types\cite{Guttman:2010io}.  Although these mapping-based methods have
been useful in detecting both splice junctions and isoforms, they rely
heavily on a high-quality reference genome.  Hence, it is not
necessarily practical to apply these methods to organisms lacking such
a reference.

The requirement for a high-quality reference genome
can be overcome by {\em de novo} assembly of short reads.
A number of {\em de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb} was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
approach is capable of increasing sensitivity of detecting
alternative isoforms over a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior to other {\em de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

@@

We separately used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While these ``local'' assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly (Table~\ref{unique_sequences},

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank (Fig.~\ref{chick_est_venn}).  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models, but not both, suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

@@

For organisms with a reference genome and Ensembl annotation, building gene
models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
reference guide seems to be a preferable method for gene and isoform expression
analysis.  However, we find that Ensembl models in chicken are missing a
substantial number of splice variants found in ESTs and mRNAs, and Cufflinks does
not recover a number of them.  A previous study by Schulz \emph{et
al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
assembly in detecting genes and isoforms in Ensembl annotations in mouse.
We find that Cufflinks almost exclusively finds splice junctions in
mouse Ensembl gene models, whereas our gene models include splice junctions that are not in
Ensembl gene models but that {\em are} well supported by ESTs (Figure~\ref{mus_venn}).

To build mouse Cufflinks gene models, we used Tophat default parameters claimed
to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
detected splice junctions not in Ensembl.  This suggests that the parameters may
have been set based on mammalian Ensembl models, which could limit the
efficiency of finding novel splice variants in other non-mammal organisms.  In
contrast, {\em de novo} assembly does not rely on species-specific parameters.
We therefore recommend that a combination of Cufflinks models and
transcripts from {\em de novo assembly} be used to build gene models that
include more splice variants.

@@

Both Cufflinks and our pipeline rely on a reference genome; therefore,
the quality of the genome will greatly affect the quality of the gene models.
In this study gene models were built from chicken genome version 2.1 (galGal3),
which contains ~17 Mb of sequence duplications and missassemblies
that were eliminated in the latest version of genome assembly
(galGal4).  Duplications and misassemblies lead to false splice junctions, which
in turn produce a large number of splice variants as observed in some
chromosomes.

@@

\subsection*{Data}

The mouse RNA-Seq dataset (SRX062280) was downloaded from the Short Read
Archive.  Chicken RNA-Seq datasets were obtained from
sequencing of mRNAs from spleen of chicken line 6 and 7.

@@

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons one of whose splice sites perfectly match intron boundaries.
Exons are considered incomplete and eliminated if they locate at the $3'$ or
$5'$ end of the transcripts and they are not the largest exons.
% @CTB ???
  ,
Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
they have at least one intron or exon in common.  Then, a splice graph composed
of exons is created and structures of isoforms are derived from traversing
paths in the splice graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

To identify unique sequences from two datasets, a set of 20-mers is created for
both datasets using khmer\cite{khmer} (@CTB what script? I may have just removed it :).  Then, 20-mers from a query dataset are
compared with 20-mers from the target dataset.  The sequence is considered
unique if more than 90\% of 20-mers in the query is unique.  Any unique region
shorter than 300 bp is ignored.
